Bayesian Inference in Numerical Cognition: A Tutorial Using JASP

### 2019-11-21 jncTutorial/manuscript.Rmd
First adaptation

author: 
  - name          : "Thomas J. Faulkenberry"
    corresponding : yes    
    affiliation   : "1"
    address       : "Department of Psychological Sciences, Box T-0820, Tarleton State University, Stephenville, TX 76401"
    email         : "faulkenberry@tarleton.edu"
  - name          : "Alexander Ly"
    affiliation   : "2"
    corresponding : no
  - name          : "Eric-Jan Wagenmakers"
    affiliation   : "2"
    corresponding : no

affiliation:
  - id            : "1"
    institution   : "Tarleton State University"
  - id            : "2"
    institution   : "University of Amsterdam"

@@

Though the basic ideas of hypothesis testing are likely familiar to anyone reading this tutorial, it may be quite instructive to take a step back and talk about what we are really doing at the most basic level, before we begin making any distinctions between frequentist and Bayesian methods. Regardless of the exact form of the hypothesis test being employed, the problem of inference reduces to a simple framework, the flow of which is displayed in Figure \@ref(fig:genModel). First, we fix a *generative* model of nature, by which we mean a statistical model that can make quantitative predictions of nature. For example, we learn in our elementary statistics courses that the normal distribution is a good model for how sample means are distributed if the sample size is large enough. Thus, we can use the normal distribution as a generative model for situations where our data is in the form of a sample mean.

![\label{fig:genModel}A diagram showing the basic problem of statistical inference. For a given inference problem, we fix a generative model that is capable of making predictions about the data we should observe under this model (solid line). We then try to invert this process -- given some observed data, we try to ascertain the true nature of this model (dashed line).](figures/genModel.pdf){width=70%}

For us, however, this flow within this modeling framework is usually inverted. We are confronted with observed data, and our problem is to ascertain the properties of the model which could have conceivably generated these data (see the dashed line of Figure \@ref(fig:genModel)). For example, consider a group of people who are given a treatment to improve their mathematical ability, which we measure using a (fictitious) scale called the Scale for Advancing Mathematical Ability (SAMA). For the general population, SAMA scores are normally distributed with a mean of 50 and a standard deviation of 10. For our participants who have received the treatment, we observe a sample mean of 55. What can we say about the population of people who have received this treatment (that is, the population from which this sample has been drawn)?

To approach this problem using hypothesis testing, we first define two competing generative models. Under one model, denoted $\mathcal{H}_0$, the SAMA scores for the treatment group are the same as in the general population (that is, 50). This model is typically called the *null model* (or *null hypothesis*), and can be written conveniently using symbols as
\[
\mathcal{H}_0:\mu=50,
\]
where $\mu$ represents the mean of the population of people who have had the treatment. For the other model $\mathcal{H}_1$, we want to capture the prediction that SAMA scores will *increase* after treatment. Thus, we define this model as 
\[
\mathcal{H}_1:\mu >50.
\]
Our problem now reduces to one simple question: which model ($\mathcal{H}_0$ or $\mathcal{H}_1$) better predicts our observed data?

From here, we have two paths we can travel. The frequentist path works roughly as follows. First, we assess how well the null hypothesis is able to predict our data. We do this by computing the probability of observing data at least as extreme as our observed sample mean of 55 *if* the null hypothesis is the true model. In symbols, we write this *conditional probability* as $p(\text{data}\mid \mathcal{H}_0)$. In the standard null hypothesis testing framework that we all learned previously, this probability is called the "$p$-value". If this $p$-value is small (i.e., $p<0.05$), we conclude that our data is rare under $\mathcal{H}_0$, and consequently, we reject $\mathcal{H}_0$ as a candidate in our list of possible generative models. Since we only defined two models, rejecting $\mathcal{H}_0$ leaves only $\mathcal{H}_1$ remaining as a candidate.

In this tutorial, we will advocate a fundamentally different approach to this problem and employ Bayesian hypothesis testing. The advantages of Bayesian hypothesis testing have been clearly outlined in a number of previous papers [@wagenmakers2007], but we will point out two here. First, we mentioned that if the $p$-value was small, we could reject $\mathcal{H}_0$, owing to the reasoning that if the null model was good at predicting our observed data, then our data should not have been so rare (we did observe it, after all). However, what would have happened if our $p$-value had been large? It turns out that this would have been completely uninformative, because it is known that under $\mathcal{H}_0$, $p$-values are uniformly distributed (REFERENCE?). So, the $p$-value gives us no indication of whether $\mathcal{H}_0$ is a good model. This leads to a problem of asymmetry; that is, while we can reject $\mathcal{H}_0$, we can never accept $\mathcal{H}_0$.

Another issue is this; recall that in the frequentist procedure described above, the $p$-value measured how well $\mathcal{H}_0$ was able to predict our observed data. However, there was no accompanying measure of the predictive adequacy of $\mathcal{H}_1$. It could be the case that while $\mathcal{H}_0$ was not a good predictive model of our observed data, $\mathcal{H}_1$ might be even worse! We simply do not know. To circumvent this problem, we will use the *Bayes factor* [@kass1995], a ratio which directly indexes the predictive adequacy of $\mathcal{H}_0$ against the predictive adequacy of $\mathcal{H}_1$. Compared to $p$-values, Bayes factors will offer a number of inviting advantages, including the ability to measure support *for* the null hypothesis $\mathcal{H}_0$, as well as the ability to simultaneously measure the ability of both $\mathcal{H}_0$ and $\mathcal{H}_1$ to predict our observed data.

## What is the Bayes factor?

To define the Bayes factor, we must first recall a basic rule of probability known as Bayes' rule:
\begin{equation}\label{eq:bayesRule}
\underbrace{p(\mathcal{M}\mid \text{data})}_{\substack{\text{Posterior beliefs}\\ \text{about model}}} = \underbrace{p(\mathcal{M})}_{\substack{\text{Prior beliefs}\\ \text{about model}}} \times \underbrace{\frac{p(\text{data}\mid \mathcal{M})}{p(\text{data})}}_{\text{predictive updating factor}}.
\end{equation}
This equation tells us that our posterior belief about a model $\mathcal{M}$ (i.e., $p(\mathcal{M}\mid \text{data})$) can be found by taking our *prior* belief about $\mathcal{M}$ and multiplying it by the fraction $p(\text{data}\mid \mathcal{M})/p(\text{data})$. This fraction is a *predictive updating factor* in the sense that it indexes how well our data is predicted by $\mathcal{M}$ compared to how well it is predicted by all other possible models. Now, since the goal here is to directly compare the predictive adequacy of two models, say $\mathcal{H}_0$ and $\mathcal{H}_1$, we can use Equation \ref{eq:bayesRule} to derive the following:
\begin{align*}
  \frac{p(\mathcal{H}_0\mid \text{data})}{p(\mathcal{H}_1\mid \text{data})}
  & = \frac{p(\mathcal{H}_0) \cdot \frac{p(\text{data}\mid \mathcal{H}_0)}{p(\text{data})}}{p(\mathcal{H}_1) \cdot  \frac{p(\text{data}\mid \mathcal{H}_1)}{p(\text{data})}} \\ 
  & = \frac{p(\mathcal{H}_0)\cdot p(\text{data}\mid \mathcal{H}_0)}{p(\mathcal{H}_1)\cdot p(\text{data}\mid \mathcal{H}_1)},\\
\end{align*}
which we can simplify to
\begin{equation}\label{eq:bayesFactor}
\underbrace{\frac{p(\mathcal{H}_0\mid \text{data})}{p(\mathcal{H}_1\mid \text{data})}}_{\substack{\text{posterior beliefs}\\ \text{about models}}} = \underbrace{\frac{p(\mathcal{H}_0)}{p(\mathcal{H}_1)}}_{\substack{\text{prior beliefs}\\ \text{about models}}} \times \underbrace{\frac{p(\text{data}\mid \mathcal{H}_0)}{p(\text{data}\mid \mathcal{H}_1)}}_{\text{predictive updating factor}}.
\end{equation}
The predictive updating factor
\[
BF_{01} = \frac{p(\text{data}\mid\mathcal{H}_0)}{p(\text{data}\mid\mathcal{H}_1)}
\]
is called the *Bayes factor*, and as we can see in Equation \@ref(eq:bayesFactor), it represents the factor by which our relative belief in $\mathcal{H}_0$ over $\mathcal{H}_1$ can be updated after observing data. Figure \@ref(fig:bayesUpdating) shows a simple visual depiction of this idea. The top line of the figure shows two observers, which we'll call Cooley and Winky (respectively, from left to right). Both observers have different prior beliefs about the relative plausibility of $\mathcal{H}_0$ and $\mathcal{H}_1$. Cooley is has slightly more belief in $\mathcal{H}_0$ than in $\mathcal{H}_1$, but is not terribly committed either way. On the other hand, Winky has a large degree of prior belief in $\mathcal{H}_1$, and consequently, less belief in $\mathcal{H}_0$. The second line shows their relative belief after observing the same data (i.e., their posterior beliefs). Smiley has moved from being quite equivocal to being solidly in favor of $\mathcal{H}_0$ after observing data. Winky has moved toward the middle, formerly having a large degree of belief in $\mathcal{H}_1$, but now having less belief in $\mathcal{H}_1$.

![\label{fig:bayesUpdating}Visual depiction of how observing data updates relative belief in two hypotheses $\mathcal{H}_0$ and $\mathcal{H}_1$.](figures/bayesUpdating.pdf){width=70%}

The important take home of Figure \@ref(fig:bayesUpdating) is this; both observers started and ended at different points on the continuum of relative belief between $\mathcal{H}_0$ and $\mathcal{H}_1$. However, *both* observers shifted the same amount\footnote{While the figure makes the shift appear as an additive shift, the shift is really multiplicative. If this discrepancy between the figure and the mathematical reality of Bayesian updating makes the reader uncomfortable, the reader may simply assume that Cooley and Winky are sitting on a logarithmic scale. They won't mind.} in the same direction (toward $\mathcal{H}_0$). Thus, these observed data are *evidential* for $\mathcal{H}_0$, and the size of the evidential shift is given by the Bayes factor.

As a concrete example, suppose we obtain a Bayes factor of $BF_{01}=12$. This says several things:

- The observed data are 12 times more likely under $\mathcal{H}_0$ than under $\mathcal{H}_1$.
- The observed data are 1/12 as likely under $\mathcal{H}_1$ compared to $\mathcal{H}_0$.
- Our prior relative belief in $\mathcal{H}_0$ over $\mathcal{H}_1$ has increased by a factor of 12.

\noindent Numerically, the Bayes factor is simply a ratio, and as such, should be compared to a baseline of 1. If $BF_{01}=1$, the data are equivocal; they are equally likely under both $\mathcal{H}_0$ and $\mathcal{H}_1$. If $BF_{01}>1$, the data support $\mathcal{H}_0$. If $BF_{01}<1$, the data support $\mathcal{H}_1$. As $BF_{01}$ increases toward $+\infty$, so does the evidence for $\mathcal{H}_0$. As $BF_{01}$ decreases down to 0, the evidence for $\mathcal{H}_1$ increases. Note also that we can cast any Bayes factor (which necessarily compares two hypotheses) in terms of the other hypothesis by taking reciprocals. That is, $BF_{10} = 1/BF_{01}$. For example, if $BF_{01} = 5$, then $BF_{10}=1/5$.

Unlike the $p$-value, there are not any universal standards for how large a Bayes factor should be before it is considered "evidential". Any Bayes factor larger than 1 is evidential for some hypothesis. However, people have made recommendations in the past regarding sizes of Bayes factors, and these will provide us with a good starting place. Particularly, we mention the recommendations of @jeffreys1961, who proposed the following scheme\footnote{We advise readers not to get too hung up on the specific labels and breakpoints in this table.}:

\begin{tabular}{rl}
Bayes factor & Evidence\\
\hline
1-3 & anecdotal\\
3-10 & moderate\\
10-30 & strong\\
30-100 & very strong\\
$>$100 & extreme\\
\end{tabular}

# Tutorial examples
In this section, we describe a number of practical examples of using Bayesian inference in the context of numerical cognition. In the examples, each based on a well-cited paper in numerical cognition, we demonstrate a number of JASP's features using  synthetic datasets, each generated in R using specific model assumptions that closely match the patterns of observed data described in the original papers.  The datasets, as well as the R script used to generate the datasets, are available for download from Github.

## Example 1: a Bayesian $t$-test

To demonstrate the Bayesian $t$-test, we will use a dataset based on @vergutsDeMoor2005, who investigated whether two-digit numbers are represented holistically on a mental number line, or instead are processed by decomposing into separate digit components. In their experiment, @vergutsDeMoor2005 presented subjects with pairs of two-digit numbers who had the same decade (e.g., 53 versus 51) or decades that differed by one (e.g., 53 versus 61).  Their reasoning for this manipulation was as follows: if subjects are using holistic processes to compare two-digit numbers, then they should exhibit a numerical distance effect [@moyer1967] on same-decade as well as different-decade pairs. That is, response times should be faster when the numerical distance between the two numbers is large, and slower when the numerical distance is small.  If, on the other hand, subjects are using decomposed processes, this numerical distance effect should disappear on the different-decade pairs. This is because in different-decade pairs, such as 53 versus 61, the comparison can be made solely on the basis of the decade digit alone. Thus, response times in this condition should be immune to manipulation of numerical distance.

In their original paper, @vergutsDeMoor2005 found support for decomposed processing on the basis of several single sample $t$-tests. For subjects asked to choose the smaller of the digit pairs, they found (as expected) a statistically significant numerical distance effect for same-decade pairs, $t(22) = -3.86$, $p<0.001$. However, the distance effect was not significantly different from 0 for different-decade pairs, $t(22)=-0.14$, $p=0.884$. A similar pattern was found for subjects who were asked to choose the larger of each pair: there was a significant distance effect for same-decade pairs, $t(19)=-5.31$, $p<0.001$, but no significant distance effect for different-decade pairs, $t(19)=0.70$, $p=0.479$.

### Theory of the Bayesian $t$-test

This section will describe how the Bayesian $t$-test works.

@@

We will now describe a Bayesian analysis inspired by the original methods of @vergutsDeMoor2005. After downloading the file \texttt{ttest.csv} from the Github repository, you can open the file in JASP by clicking the menu button -- "Open" -- "Computer", which will then open a file browser for you to locate the downloaded file on your computer. Once the file is opened, you will be able to see a dataset with three variables (see Figure \@ref(fig:ttestData). The variable \texttt{condition} defines whether each subject was instructed to choose the smaller or the larger of each presented two-digit number pair. The variable \texttt{distSame} represents each subject's slope after regressing response times against numerical distance [c.f., @lorch1990] for same-decade pairs. Note that a negative slope is indicative of the numerical distance effect, as response times decrease as a function of increasing numerical distance. Finally, \texttt{distDifferent} represents each subject's regression slope for different-decade pairs.

@@

We are now ready to perform a Bayesian analysis of the simulated @vergutsDeMoor2005 data. First, since we will separately analyze the "Choose smaller" and "Choose larger" groups, we will need to apply a *filter* to the dataset; specifically, we will first consider only the "Choose smaller" group. This is done in JASP by clicking on the filter icon (a small, funnel-shaped icon in the top left corner of the data sheet). After clicking on the funnel, a graphical drag-and-drop interface will appear at the top of the window. Since we are filtering on the variable \texttt{condition}, simply click the variable name on the left side of the filter dialogue.  This will make the variable name appear in the dialogue. Then, we click on the "=" symbol at the top of the dialogue. After this, type the name of the level we wish to retain; that is, we type \texttt{smaller} (without quotes). Finally, click the button at the bottom of the dialogue that says "Apply pass-through filter". When successful, the words will change to "Filter applied", and the bottom of the JASP window will update to tell us how many of the data rows passed through the filter (in this case, 23 will pass). You will also see that any row where \texttt{condition} is equal to "larger" will be grayed out.

To perform a Bayesian one-sample $t$-test, click on the "T-Tests" button and select "Bayesian One Sample T-Test". There are two $t$-tests that we will wish to perform, but for clarity, we'll start with describing one in detail. Before beginning, however, we will define our two competing hypotheses explicitly. Since the numerical distance effect is expected to be negative, we will define $\mathcal{H}_0:\delta = 0$ and $\mathcal{H}_1:\delta < 0$.

We will first perform a $t$-test for the same-decade trials. From the variable list, move "distSame" over to the "Variables" box. Also, our alternative hypothesis defined so that the effect is negative, so we will choose "$<$ Test value" from the "Alt. Hypothesis" menu. Finally, select also "Prior and posterior" from the "Plots" menu. This setup and the resulting output appears in Figure \@ref(fig:ttestBayes).  There are two components of this output that we will now discuss in detail: the computed Bayes factor (top of the Results pane in Figure \@ref(fig:ttestBayes)) and the plot of the prior and posterior for the effect size parameter $\delta$ (bottom of the Results pane in Figure \@ref(fig:ttestBayes)).

![\label{fig:ttestBayes}JASP screenshot showing the Bayesian one-sample $t$-test for \texttt{distSame}.](figures/ttestBayes.png){width=90%}

>>>>

author: 
  - name          : "Thomas J. Faulkenberry"
    corresponding : yes    
    affiliation   : "1"
    address       : "Department of Psychological Sciences, Box T-0820, Tarleton State University, Stephenville, TX 76401"
    email         : "faulkenberry@tarleton.edu"
  - name          : "Alexander Ly"
    affiliation   : "2,3"
    corresponding : no
  - name          : "Eric-Jan Wagenmakers"
    affiliation   : "2"
    corresponding : no

affiliation:
  - id            : "1"
    institution   : "Tarleton State University"
  - id            : "2"
    institution   : "University of Amsterdam"
  - id            : "3"
    institution   : "Centrum Wiskunde & Informatica"

@@

<!-- To elaborate on the concept of (Bayesian) hypothesis testing,  -->

Even though the reader of this paper is probably familiar with hypothesis testing, we believe that it is nonetheless instructive to go over the basics.

\subsection{Hypotheses}
Hypothesis testing can be viewed as validating general claims based on a small sample of the general population. For illustrative purposes, suppose that we are policy makers confronted with a new study programme, which claims to increase the mathematical abilities of sixth graders in the United States. To ease the exposition we use a (fictitious) scale called the Scale for Advancing Mathematical Ability (SAMA) to measure mathematical ability. As policy makers we know that the population mean SAMA score is \( \mu_{0}=50 \) and that the standard deviation is \( \sigma=10 \) for the population of sixth graders that are taught using the standard materials in the US.

The superiority claim about the new study programme can be formalised as the working hypothesis \( \mathcal{H}_{1} : \mu > 50 \). Here \( \mu \) refers to the \emph{population} mean of SAMA score of the same sixth graders, but taught using the new study programme instead. Such a treatment population does not (yet) exist and \( \mu \) cannot be measured, and is, therefore, unknown.

To nonetheless infer the validity of the working hypothesis \( \mathcal{H}_{1} \), we run an experiment in which \( n=90 \) sixth graders are taught with the new study programme. Suppose that these particular participants scored an average SAMA score of \( \bar{x} = 52.07 \). Clearly, this sample mean is larger than the status quo population mean of \( 50 \), but can we assume that this finding generalises to the whole population? If the answer is "yes" and we declare the working hypothesis to be verified, then how confident would we be about such a decision? Below we recall how \( p \)-values are typically used to make hard "yes" and "no" decisions, and we remind the reader why a \( p \)-value does not quantify the (un)certainty of such decisions.

\subsection{Statistical models}
To deduce how hypotheses, claims about the population such as \( \mu \), affect the data, such as \( \bar{x} \), we have to link the former to the latter with a distribution, for instance, with a normal distribution. The population parameters, the distribution that links them to the data, and the data together make up a (statistical) model, which allows for quantitative predictions. For instance, the claim that the treatment population is identitical to the current population of sixth graders implies the null model \( \mathcal{M}_{0} \) comprised of (1) the null hypothesis \( \mathcal{H}_{0}: \mu = 50 \), (2) the assumption \( \sigma = 10 \), and (3) normally distributed SAMA scores. When all three assumptions hold true, we then deduce that the \emph{potential} average SAMA score consisting of a group of \( n = 90 \) participants is expected to fall within the interval \( 47.93 = 50 - 1.96 \frac{10}{\sqrt{n}} \) and \( 52.07 = 50 + 1.96 \frac{10}{\sqrt{n}} \) with 95% chance. <!-- 
-->
\footnote{We use ``chance'' to refer to probabilistic statements about the data.}
<!-- 
-->

\subsection{Null hypothesis significance testing}
Reasoning from the general population under the assumption that the parameters are known to derive statements about potential outcomes forms the basis of null hypothesis signficance testing. By definition the \( p \)-value is the chance of seeing the observed outcome and more extreme --but \emph{unobserved}-- outcomes under the null.

Under the null model \( \mathcal{M}_{0} \), the more extreme data are potential outcomes of \( \bar{x} \) larger than the observed sample mean of \( 52.07 \), but also lower than the unobserved possible sample mean of \( 47.93 = 50 - (52.07-50) \). Under the null, and the normality assumptions in particular, allows us to deduce that the chance of seeing a potential \( \bar{x} \) in \( (- \infty, 47.93] \) or \( [52.07, \infty) \) is 5%, thus, \( p=0.05 \). Note that \( 1 - p = 0.95 \), which is the 95% chance of the non-extreme event of having \( \bar{x} \) fall within the interval \( [47.93, 52.07] \), if the null model holds true. We use this example to emphasise that \( p \)-values help us reason from the null model to (potential) data, but they do not allow us to make quantitative statements from the actually observed data about the hypotheses.

A low \( p \)-value has many causes, and for the case we mention four: (1) the null hypothesis is not true, (2) the true variance is actually higher than \( 10 \), (3) the data are not normally distributed, or (4) the null model actually holds true, but we saw a rare event. Despite the various causes of a low \( p \), typically only (1) is singled out and mentioned as \emph{the} culprit. The typical conclusion is then to reject the null, and the working hypothesis \( \mathcal{H}_{1} :  \mu > 50 \) is proposed as the winner. Note that this type of reasoning provides only an indirect argument for our working hypothesis, as the predictions of the working hypothesis itself were not considered at all. Only the null was used in the computation of \( p \). The fact that the null is not a good fit for the data, does not automatically imply that the alternative fit the data well.

We conclude our summier review on \( p \)-values with three observations: (i) the comparison between \( \mathcal{H}_{0} \) and \( \mathcal{H}_{1} \) is unfair, (ii) \( p \)-values only allow us to deduce predictions from the general population to potential data, but not the other way around, and (iii) \( p \)-values can only lead to hard reject or non-reject decisions, which ignores the nuances of a scientific enquiry.

<!--
the \( p \)-value does not provide us with a measure of uncertainty about the hypotheses, because they are chance statements about potential data, and not about the hypotheses.

The latter implies that one can only come to the conclusion of accept or reject, without  -->

\subsection{Bayesian testing}
Bayesian testing addresses these problems as follows: Firstly, it compares the two hypotheses on equal footing by also  operationalises the alternative model. For the case at hand, \( \mathcal{M}_{1} \) would be comprised of the assumptions (1') \( \mathcal{H}_{1}: \mu > 50 \), (2) \( \sigma =10 \) as before, and (3) the data being normally distributed as before. By doing so, violation of normality and the assumed standard deviation will hit both models equally. Secondly, to reason from a particular observed outcome to the general population, it employs Bayes rule. To do so, we require to express our uncertainty about the two hypotheses in terms of so-called prior model probabilities before data observation.

Prior model probabilities help us contextualise the testing problem. Working as policy makers, we might have seen many claims of improved study programmes that did not work out well. For instance, if 4-to-1 claims of improved study programmes led to null results, we might set \( P( \mathcal{M}_{1}) = 0.2 \) and \( P(\mathcal{M}_{0}) = 0.8 \). On the other hand, the designers of the new study programme to increase SAMA scores might choose \( P( \mathcal{M}_{1}) = 0.7 \) and \( P(\mathcal{M}_{0}) = 0.3 \) instead. If \( \mathcal{M}_{1} \) and \( \mathcal{M}_{0} \) are the only models under consideration we have \( 0 < P( \mathcal{M}_{0}) < 1 \) and \( P( \mathcal{M}_{1})= 1 - P( \mathcal{M}_{0}) \). The Bayes factor \( \textnormal{BF}_{10}(d) \) is a function of the observed data \( d \) that updates these prior model probabilities to posterior model probabilities, as follows:

\begin{align}
\label{eqPriorToPosterior}
P(\mathcal{M}_{1} \, | \, d) = \frac{ \textnormal{BF}_{10}(d) P(\mathcal{M}_{1})}{\textnormal{BF}_{10}(d) P(\mathcal{M}_{1}) + P(\mathcal{M}_{0})} \text{, and } P(\mathcal{M}_{0} \, | \, d) = 1 - P(\mathcal{M}_{1} \, | \, d). 
\end{align}

Note that the data only enter the formula via the Bayes factor. For instance, if \( \textnormal{BF}_{10}(d) = 6 \), then the sceptic's probability of the superiority of the new study programme increases to 60%, i.e., \( P(\mathcal{M}_{1} \, | \, d)=0.60 \), and retains 40% probability of for the null after data observation. On the other hand, the proponent of the new study programme would report a posterior model probability of \( P(\mathcal{M}_{1} \, | \, d)=0.93 \) and \( P(\mathcal{M}_{0} \, | \, d)=0.07 \). Note that the posterior model probabilities combine what we believed prior to data observation with the evidence of the data in terms of the Bayes factors. Both sceptic and the proponent could, if they wish, declare the new study programme to be superior to the status quo, but the sceptic would be much less confident than the proponent. The resulting posterior model probabilities can, thus, be used to complement a decision with a measure of uncertainty.

**Bayes factor interpretations**
Prior model probabilities are context dependent, whereas Bayes factors only rely on the statistical design. To simplify notation we drop the reference to the data in the Bayes factor and denote them as \( \textnormal{BF}_{10} \), where the 1 in the subscript refers to the alternative model, and the 0 to the null model. Bayes factor ranges from zero to \( + \infty \), and \( \textnormal{BF}_{10} = 8 \) implies that the observed data are 8 times more likely under the alternative compared to the null model. The reciprocal of Bayes factors smaller than one are typically easier to interpret. For instance, \( \textnormal{BF}_{10} = 0.2 \) is easier interpret as \( \textnormal{BF}_{01} = 1/\textnormal{BF}_{10} = 5 \), which means that the observed data are 5 times more likely under the null compared to the alternative. Bayes factors are best viewed as a continuous measure of evidence. In general, \( \textnormal{BF}_{10} \) larger than one are perceived to provide evidence for the alternative against the null and if it grows to \( \infty \), the posterior model probability \( P(\mathcal{M}_{1} \, | \, d) \) grows to one, thus the more certain we will be about the adequacy of the alternative.

Unlike the \( p \)-value, there are no universal standards for when a Bayes factor provides a sufficient amount of evidence. However, people have made recommendations in the past regarding sizes of Bayes factors, and these provide us with a helpful, but very rough guideline on its interpretation, when we do not know the context of the problem. In particular, we mention the recommendations of @jeffreys1961, who proposed the following scheme\footnote{We advise readers not to get too hung up on the specific labels and breakpoints in this table, see also @giollaInpresswhat.}:

\begin{tabular}{rl}
Bayes factor & Evidence\\
\hline
1-3 & anecdotal\\
3-10 & moderate\\
10-30 & strong\\
30-100 & very strong\\
\( > \)100 & extreme\\
\end{tabular}

In principle, the Bayes factor does not come with a specified decision rule. We believe that context made explicit using prior model probabilities by field experts are much more important, than overly generalised "rules" such \( p < 0.05 \) or \( \textnormal{BF} _{10} > k \), where \( k \) is some arbitrarily chosen number.

# Tutorial examples
In this section, we describe a number of practical examples of using Bayesian inference in the context of numerical cognition. In the examples, each based on a well-cited paper in numerical cognition, we demonstrate a number of JASP's features using synthetic data sets, each generated in R using specific model assumptions that closely match the patterns of observed data described in the original papers.  The datasets, as well as the R script used to generate the datasets, are available for download from Github.

## Example 1: a Bayesian \( t \)-test

To demonstrate the Bayesian \( t \)-test, we will use a dataset based on @vergutsDeMoor2005, who investigated whether two-digit numbers are represented holistically on a mental number line, or instead are processed by decomposing into separate digit components. In their experiment, @vergutsDeMoor2005 presented subjects with pairs of two-digit numbers who had the same decade (e.g., 53 versus 51) or decades that differed by one (e.g., 53 versus 61).  Their reasoning for this manipulation was as follows: if subjects are using holistic processes to compare two-digit numbers, then they should exhibit a numerical distance effect [@moyer1967] on same-decade as well as different-decade pairs. That is, response times should be faster when the numerical distance between the two numbers is large, and slower when the numerical distance is small.  If, on the other hand, subjects are using decomposed processes, this numerical distance effect should disappear on the different-decade pairs. This is because in different-decade pairs, such as 53 versus 61, the comparison can be made solely on the basis of the decade digit alone. Thus, response times in this condition should be immune to manipulation of numerical distance.

In their original paper, @vergutsDeMoor2005 found support for decomposed processing on the basis of several single sample \( t \)-tests. For subjects asked to choose the smaller of the digit pairs, they found (as expected) a statistically significant numerical distance effect for same-decade pairs, \( t(22) = -3.86 \), \( p<0.001 \). However, the distance effect was not significantly different from 0 for different-decade pairs, \( t(22)=-0.14 \), \( p=0.884 \). A similar pattern was found for subjects who were asked to choose the larger of each pair: there was a significant distance effect for same-decade pairs, \( t(19)=-5.31 \), \( p<0.001 \), but no significant distance effect for different-decade pairs, \( t(19)=0.70 \), \( p=0.479 \).

### Theory of the Bayesian \( t \)-test

The Bayesian \( t \)-test forms the basis for the other tests described in this paper. The technical details are included for completeness, but are not necessary for the use of Bayes factors.

Recall that a Bayes factor compares two models, say, \( \mathcal{M}_{1} \) and \( \mathcal{M}_{0} \), and suppose that the vector of free parameters of these models are denoted by \( \theta_{1} \) and \( \theta_{0} \) respectively. For instance, the \( \theta_{1} = ( \mu, \sigma) \) and \( \theta_{0} = \sigma \) for the one-sample \( t \)-test.

To construct a Bayes factor, the statistician needs to select two priors --one for each model, which are denoted by \( \pi_{1} (\theta_{1}) \) and \( \pi_{0} (\theta_{0}) \) respectively (e.g., @bayarri2012criteria), and provide its users with a computational method. We only introduced the notation \( \pi_{1} (\theta_{1}) \) and \( \pi_{0} (\theta_{0}) \), chosen by the statistician, to highlight their difference to prior model probabilities which are denoted with a capital \( P \), chosen by the field experts.

The Bayesian \( t \)-test is based on the priors \( \pi_{1} \), which puts a "uninformative" prior on the nuisance parameter \( \sigma \) (e.g., @ly2017tutorial) in conjunction with a Cauchy prior on the population effect size \( \delta = \mu / \sigma \), and the prior \( \pi_{0} \) with the same "uninformative" prior on the nuisance parameter \( \sigma \). With these priors, the resulting Bayes factor has a number of desirable properties: (1) The Bayes factor provides the same evidence independent on the units of measurements used for the data (e.g., @rouder2009bayesian). For instance, when the data consists of temperature measurements, the Bayes factor is the same whether the data were entered as Celsius, or Fahrenheit. (2) The Bayes factor is predictively matched, which means that it remains one, whenever the data set is not large enough. A Bayes factor of one implies that the posterior model probabilities equal the prior model probabilities, and nothing has been learned, see @jeffreys1961, for further details. (3) The Bayes factor is information consistent, which means that if for fixed sample size \( n \) the observed \( t \)-value goes to infinity, that the Bayes factor also goes to infinity, see @gronauInpressinformed, @ly2016harold for further details. (4) The Bayes factor is consistent, in the sense that as the data generated from the null model increases, thus, \( n \) grows, the Bayes factor \( \textnormal{BF}_{10} \) tends to zero. On the other hand, if \( n \) grows with data generated under the alternative, the Bayes factor \( \textnormal{BF}_{10} \) grows unbounded, see for instance @liang2008mixtures. (5) With the uninformative prior on \( \sigma \), this Bayes factor allows for optional stopping, that is, early stopping when there is enough evidence for the alternative (e.g., @hendriksen2018optional). On top of it also allows for optional continuation, which implies that one can build on top of previous results, see for instance @faulkenberry2019estimating, @grunwald2019safe, @ly2017replication.

With these priors at hand, the Bayes factor is defined as a ratio of marginal likelihoods, that is,

\begin{align}
\textnormal{BF}_{10} = \frac{ \int_{\Theta_{1}} f(d \, | \, \theta_{1}) \pi_{1}(\theta_{1}) \textnormal{d} \theta_{1}}{\int_{\Theta_{0}} f(d \, | \, \theta_{0}) \pi_{0}(\theta_{0}) \textnormal{d} \theta_{0}}.
\end{align}

These technical details do not matter for most users, as the computations are done by JASP in the background.

@@

We will now describe a Bayesian analysis inspired by the original methods of @vergutsDeMoor2005. After downloading the file \texttt{ttest.csv} from the Github repository, you can open the file in JASP by clicking the menu button -- "Open" -- "Computer", which will then open a file browser for you to locate the downloaded file on your computer. Once the file is opened, you will be able to see a dataset with three variables (see Figure \@ref(fig:ttestData). The variable \texttt{condition} defines whether a participant was instructed to choose the smaller or the larger of each presented two-digit number pair. The variable \texttt{distSame} represents each subject's slope after regressing response times against numerical distance [c.f., @lorch1990] for same-decade pairs. Note that a negative slope is indicative of the numerical distance effect, as response times decrease as a function of increasing numerical distance. Finally, \texttt{distDifferent} represents each subject's regression slope for different-decade pairs.

@@

We are now ready to perform a Bayesian analysis of the simulated @vergutsDeMoor2005 data. First, since we will separately analyze the "Choose smaller" and "Choose larger" groups, we will need to apply a *filter* to the dataset; specifically, we will first consider only the "Choose smaller" group. <!--@Tom, Would it be easier if we explained this using the click filter -->This is done in JASP by clicking on the filter icon (a small, funnel-shaped icon in the top left corner of the data sheet). After clicking on the funnel, a graphical drag-and-drop interface will appear at the top of the window. Since we are filtering on the variable \texttt{condition}, simply click the variable name on the left side of the filter dialogue.  This will make the variable name appear in the dialogue. Then, we click on the "=" symbol at the top of the dialogue. After this, type the name of the level we wish to retain; that is, we type \texttt{smaller} (without quotes). Finally, click the button at the bottom of the dialogue that says "Apply pass-through filter". When successful, the words will change to "Filter applied", and the bottom of the JASP window will update to tell us how many of the data rows passed through the filter (in this case, 23 will pass). You will also see that any row where \texttt{condition} is equal to "larger" will be grayed out.

To perform a Bayesian one-sample \( t \)-test, click on the "T-Tests" button in the ribbon and select "Bayesian One Sample T-Test". There are two \( t \)-tests that we will wish to perform, but for clarity, we'll start with describing one in detail. Before beginning, however, we will define our two competing hypotheses explicitly. Since the numerical distance effect is expected to be negative, we will define \( \mathcal{H}_{0}:\delta = 0 \) and \( \mathcal{H}_{-}:\delta < 0 \), where the minus refers to the working hypothesis of interest.

We will first perform a \( t \)-test for the same-decade trials. From the variable list, move "distSame" over to the "Variables" box. Also, our alternative hypothesis defined so that the effect is negative, so we will choose "\( < \) Test value" from the "Alt. Hypothesis" menu. Finally, select also "Prior and posterior" from the "Plots" menu. This setup and the resulting output appears in Figure \@ref(fig:ttestBayes).  There are two components of this output that we will now discuss in detail: the computed Bayes factor (top of the Results pane in Figure \@ref(fig:ttestBayes)) and the plot of the prior and posterior for the effect size parameter \( \delta \) (bottom of the Results pane in Figure \@ref(fig:ttestBayes)).

![\label{fig:ttestBayes}JASP screenshot showing the Bayesian one-sample \( t \)-test for \texttt{distSame}.](figures/ttestBayes.png){width=90%}
